---
title: "Questions on exponential growth"
author: "Elizabeth A. Babcock"
format: html
editor: visual
---

```{r,message=FALSE}
library(tidyverse)
```

### Neal: Introduction to Population Biology (question 4.1).

Each individual of an annual plant species produces on average ten seeds. (a) Starting with a single individual, what will be the population size after 6 years, assuming all the seeds survive to produce adult plants the following generation? (b) A new variety of the plant produces 5 per cent more seeds, i.e. 10.5 seeds on average. What will be the potential number of offspring after 6 years from a single individual?

for plant 1 $N_6=10^6$

for plant 2 $N_6=10.5^6=1.34x 10^6$

The new plant has 34% more offspring in 6 years.

Make a plot of the two trends.

```{r}
plants<-data.frame(t=1:6)
plants$N1<-10^plants$t
plants$N2<-10.5^plants$t
ggplot(plants,aes(x=t))+
  geom_line(aes(y=N1,color="R=10"))+
  geom_line(aes(y=N2,color="R=10.5"))
```

### Neal: Introduction to Population Biology (question 4.2).

\(a\) The Norway rat breeds continuously and has an estimated rate of natural increase (rm) of 0.0143 per day. A small number invade a garbage dump where living conditions are ideal. How long will it take the population to double in size?

$$r_m=0.0143$$ $$N_0=1$$ $$N_t=2$$ $$t=?$$

$$N_t=N_0 exp(r_m t)$$ $$log(N_t)=r_m t$$ $$t=log(N_t)/r_m$$ $$t=log(2)/0.0143$$

```{r}
log(2)/0.0143
```

48 days for the population to double.

\(c\) How long will it take for the human population to multiply by 10?

```{r}
rm<-0.001095043
log(10)/rm
```

Population would multiply by 10 in 2103 years.

### Neal: 4.5. A population increases fivefold over a 4-week period. What is the value of $\lambda$ per day, assuming exponential growth?

```{r}
N0<-1
N4<-5
rm<-log(5)/(4*7)
rm
lambda<-exp(rm)
lambda
```

Lambda=1.06 per week.
