---
title: '1106'
author: "Sierra Hamlin"
date: "2025-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this question, we will use the obsdat dataset, but with the swordfish column (SWO) rather than BUM.  
```{r}
library(tidyverse)
library(glmmTMB)
library(MuMIn)
library(DHARMa)
library(mgcv)
theme_set(theme_bw())
```


1.Fit a Poisson model with your choice of predictor variables, along with the sets offset, and look at the DHARMa residuals. Does it fit well? NO, it does not fit well.
```{r}
obsdat <- read.csv("obsdatSwo.csv")%>%
 mutate(Year=factor(Year),
        fleet=factor(fleet),
        season=factor(season),
        area=factor(area))
ggplot(obsdat,
       aes(x=Year,y=SWO))+
  stat_summary(fun.data="mean_se")
ggplot(obsdat,aes(x=SWO))+
  geom_histogram(binwidth=1)

poissonFit<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="poisson")
summary(poissonFit)
plot(simulateResiduals(poissonFit))
```

2. Fit a negative binomial model and look at the DHARMa residuals. Does it fit well?  Better than poisson but not great.
```{r}
NB2fit<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2")
summary(NB2fit)
plot(simulateResiduals(NB2fit))
```


3. Use the dredge function to find the BIC best model starting the model with all two way interactions. Which model is best? Model 66 is the best fit.
```{r}
NB2fit<-glmmTMB(SWO~(Year+area+season)^2+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2",
                na.action = "na.fail")

NB2dredge<-dredge(NB2fit,rank = "BIC")
head(NB2dredge)
NB2best<-get.models(NB2dredge,subset=1)[[1]]
summary(NB2best)
```







