---
title: "Catch, effort and CPUE"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r,message=FALSE}
library(tidyverse)
library(glmmTMB)
library(MuMIn)
library(DHARMa)
library(mgcv)
theme_set(theme_bw())

```

```{r}
obsdat<-read.csv("obsdat.csv") %>%
 mutate(Year=factor(Year),
        fleet=factor(fleet),
        season=factor(season),
        area=factor(area))
summary(obsdat)
ggplot(obsdat,
       aes(x=Year,y=BUM))+
  stat_summary(fun.data="mean_se")
ggplot(obsdat,aes(x=BUM))+
  geom_histogram(binwidth=1)
```


Poisson vs. nb 1 or 2

```{r}
df<-expand.grid(draw=1:1000,model=c("Poisson","Negative Binomial")) %>%
  rowwise()%>%
  mutate(Draw=ifelse(model=="Poisson",rpois(1,10),rnbinom(1,mu=10,size=0.5)))
summary(df)
df%>% group_by(model)%>%
  summarize(mean=mean(Draw),
            var=var(Draw))
ggplot(df,aes(x=Draw,fill=model))+
  geom_histogram(position="dodge")
```


Run a poisson model and look at the residuals

```{r}
poissonFit<-glmmTMB(BUM~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="poisson")
summary(poissonFit)
plot(simulateResiduals(poissonFit))
```

Poisson doesn't work, try NB2

```{r}
NB2fit<-glmmTMB(BUM~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2")
summary(NB2fit)
plot(simulateResiduals(NB2fit))

```

Looks fine.


```{r}
NB1fit<-glmmTMB(BUM~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom1")
plot(simulateResiduals(NB1fit))

```

stick with NB2 and find best predictors using dredge

```{r,message=FALSE}
NB2fit<-glmmTMB(BUM~(Year+area+season)^2+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2",
                na.action = "na.fail")
#Must set na.action to "na.fail"
#(Year+area+season)^2
#Year+area+season+Year:area+Year:season+area:season

NB2dredge<-dredge(NB2fit,rank = "BIC")
NB2dredge2<-dredge(NB2fit)
head(NB2dredge2)
NB2best<-get.models(NB2dredge,subset=1)[[1]]
#summary(NB2best)
```
Best model includes all main effects, plus interactions (but not with year)

Prediction, for N and S separately

```{r}
newdata<-expand.grid(Year=factor(levels(obsdat$Year)),
                    area=factor(levels(obsdat$area)),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(NB2best,
              newdata = newdata,
              se.fit = TRUE,
              type="response")
newdata$fit<-temp$fit
newdata$se<-temp$se.fit
ggplot(newdata,aes(x=as.numeric(as.character(Year)),
                   y=fit,ymin=fit-se,ymax=fit+se,color=area))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)",
       main="Negative binomial fit")
  
```


Tweedie

```{r}

TweedieFit<-glmmTMB(BUM/sets~(Year+area+season)^2,
                  data=obsdat,
                  family="tweedie",
                na.action = "na.fail")
#Must set na.action to "na.fail"

TweedieDredge<-dredge(TweedieFit,rank = "BIC")
head(TweedieDredge)
TweedieBest<-get.models(TweedieDredge,subset=1)[[1]]
summary(TweedieBest)
plot(simulateResiduals(TweedieBest))
```

Residuals look fine. Prediction

```{r}
newdata2<-expand.grid(Year=factor(levels(obsdat$Year)),
                    area=factor("N"),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(TweedieBest,
              newdata = newdata2,
              se.fit = TRUE,
              type="response")
newdata2$fit<-temp$fit
newdata2$se<-temp$se.fit
bind_rows(list(NB2=filter(newdata,area=="N"),
          Tweedie=newdata2),.id="type") %>%
ggplot(aes(x=as.numeric(as.character(Year)),y=fit,ymin=fit-se,
           ymax=fit+se,color=type))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)")

```
Delta-lognormal model

Will need the lognormal conversion functions. These calculate the mean and standard error of a lognormal function from the mean and standard error of the normally distributed variable on the log scale. Also need the function for Lo et al. to calculate the standard error of the index using the delta method. 

```{r}
lnorm.mean=function(x1,x1e) {
  exp(x1+0.5*x1e^2)
}

lnorm.se=function(x1,x1e) {
  ((exp(x1e^2)-1)*exp(2*x1+x1e^2))^0.5
}
lo.se=function(x1,x1e,x2,x2e) {
  if(length(x1[!is.na(x1) &!is.na(x2)])>1 & length(unique(x1))>1 & length(unique(x2))>1 )
   cor12=cor(x1[!is.na(x1) &!is.na(x2)],x2[!is.na(x1) &!is.na(x2)]) else cor12=0
  (x1e^2 * x2^2 +x2e^2*x1^2+2*x1*x2*cor12*x1e*x2e)^0.5
}

```


Part 1. binomial


```{r}
obsdat$present<-ifelse(obsdat$BUM>0,1,0)
BinomFit<-glmmTMB(present~Year+area*season,
                  data=obsdat,
                  family="binomial",
                na.action = "na.fail")
BinDredge<-dredge(BinomFit,rank = "BIC")
head(BinDredge)
BinBest<-get.models(BinDredge,subset=1)[[1]]
summary(BinBest)
plot(simulateResiduals(BinBest))

```

Part 2. log(CPUE) positive only

```{r}
posdat<-filter(obsdat,present==1)%>%
  mutate(cpue=BUM/sets)
cpueFit<-glmmTMB(log(cpue)~Year+area*season,
                  data=posdat,
                  family="gaussian",
                na.action = "na.fail")
cpueDredge<-dredge(cpueFit,rank = "BIC")
head(cpueDredge)
cpueBest<-get.models(cpueDredge,subset=1)[[1]]
summary(cpueBest)
plot(simulateResiduals(cpueBest))

```

Calculate index

```{r}
newdata3<-expand.grid(Year=factor(levels(obsdat$Year)),
                    area=factor("N"),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(BinBest,
              newdata = newdata3,
              se.fit = TRUE,
              type="response")
newdata3$binomial<-temp$fit
newdata3$binomial.se<-temp$se.fit
temp<-predict(cpueBest,
              newdata = newdata3,
              se.fit = TRUE) #response is on log scale
#lognormal conversion functions to get mean if positive
newdata3$pos.cpue<-lnorm.mean(temp$fit,temp$se.fit)
newdata3$pos.cpue.se<-lnorm.se(temp$fit,temp$se.fit)
#index
newdata3<-newdata3 %>%
  mutate(fit=binomial*pos.cpue,
         se=lo.se(binomial,binomial.se,pos.cpue,pos.cpue.se))
head(newdata3)
```

Compare fits

```{r}
bind_rows(list(NB2=filter(newdata,area=="N"),
          Tweedie=newdata2,
          DeltaLN=newdata3),.id="type") %>%
ggplot(aes(x=as.numeric(as.character(Year)),y=fit,ymin=fit-se,
           ymax=fit+se,color=type))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)")
```
Lognormal is a little flatter. 

Random effect

```{r}

NB2Random<-glmmTMB(BUM~Year+area+season+(1|area:Year)+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2")

summary(NB2Random)
```


### Gam

GAMs can be fit with the mgcv library

```{r}
NB2Gam<-gam(BUM~Year+season*area+te(lon,lat),data=obsdat,
            family="nb") #Negative binomial 2
gam.check(NB2Gam)
summary(NB2Gam)
plot(NB2Gam)

```

### Bycatch estimation

BycatchEstimator tool

https://ebabcock.github.io/BycatchEstimator/

```{r}
library(BycatchEstimator)
library(MuMIn)
setupObj<-bycatchSetup(
  obsdat = droplevels(LLSIM_BUM_Example_observer[LLSIM_BUM_Example_observer$Year>2010 &LLSIM_BUM_Example_observer$fleet==2,]),
  logdat = droplevels(LLSIM_BUM_Example_logbook[LLSIM_BUM_Example_logbook$Year>2010 & LLSIM_BUM_Example_logbook$fleet==2,]),
  yearVar = "Year",
  obsEffort = "hooks",
  logEffort = "hooks",
  obsCatch = c("SWO","BUM")[2], # selecting Blue marlin catch
  catchUnit = "number",
  catchType = "catch",
  logNum = NA,
  sampleUnit = "trips",
  factorVariables = c("Year","area","season"),
  numericVariables = NA,
  EstimateBycatch = TRUE,
  baseDir = getwd(),
  runName = "LLSIMBUMtripExample",
  runDescription = "LLSIMBUM by trip",
  common = c("Swordfish","Blue marlin")[2], # selecting Blue marlin
  sp = c("Xiphias gladius","Makaira nigricans")[2], # selecting Blue marlin
  reportType = "html"
)
designObj <- bycatchDesign(
  setupObj = setupObj,
  designScenario = "noPool",
  designMethods = c("Ratio", "Delta"),
  designVars = c("Year","area","season"),
  designPooling = FALSE,
  poolTypes=c("adjacent","all","none"),
  pooledVar=c(NA,NA),
  adjacentNum=c(1,NA),
  minStrataUnit = 1,
  reportType="html"
)
bycatchFit(
  setupObj = setupObj,
  modelScenario = "s1",
  complexModel = formula(y~Year+area),
  simpleModel = formula(y~Year),
  indexModel = formula(y~Year),
  modelTry = c("TMBnbinom1","TMBtweedie"),
  randomEffects = NULL,
  randomEffects2 = NULL,
  selectCriteria = "BIC",
  DoCrossValidation = TRUE,
  CIval = 0.05,
  VarCalc = "Simulate",
  useParallel = TRUE,
  nSims = 100,
  plotValidation = FALSE,
  trueVals = NULL,
  trueCols = NULL,
  reportType = "html"
)
allResults<-loadOutputs(baseDir = getwd(),
                      runName= "LLSIMBUMtripExample",
                      runDate =  Sys.Date(),
                      designScenarios = "noPool",
                      modelScenarios = "s1"
)
#Plot all together
ggplot(allResults$allYearEstimates,aes(x=Year,y=Total,
                                       ymin=TotalLCI,ymax=TotalUCI,
                      fill=Source,color=Source))+
  geom_line()+
  geom_ribbon(alpha=0.4)+
  theme_bw()
```


