
```{r,message=FALSE}
library(tidyverse)
library(glmmTMB)
library(MuMIn)
library(DHARMa)
library(mgcv)
theme_set(theme_bw())

```

```{r}
obsdat<-read.csv("obsdatSwo.csv") %>%
 mutate(Year=factor(Year),
        fleet=factor(fleet),
        season=factor(season),
        area=factor(area))
ggplot(obsdat,
       aes(x=Year,y=SWO))+
  stat_summary(fun.data="mean_se")
ggplot(obsdat,aes(x=SWO))+
  geom_histogram(binwidth=1)
```


Poisson vs. nb 1 or 2

```{r}
df<-expand.grid(draw=1:1000,model=c("Poisson","Negative Binomial")) %>%
  rowwise()%>%
  mutate(Draw=ifelse(model=="Poisson",rpois(1,10),rnbinom(1,mu=10,size=0.5)))
summary(df)
df%>% group_by(model)%>%
  summarize(mean=mean(Draw),
            var=var(Draw))
ggplot(df,aes(x=Draw,fill=model))+
  geom_histogram(position="dodge")
```

Run a poisson model and look at the residuals

```{r}
poissonFit<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="poisson")
summary(poissonFit)
plot(simulateResiduals(poissonFit))
```

1. Poisson does not fit well.

```{r}
NB2fit<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2")
summary(NB2fit)
plot(simulateResiduals(NB2fit))

```


```{r}
NB1fit<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom1")
plot(simulateResiduals(NB1fit))

```

None are great, but NB2 fits best.

```{r,message=FALSE}
NB2fit<-glmmTMB(SWO~(Year+area+season)^2+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2",
                na.action = "na.fail")
#Must set na.action to "na.fail"

NB2dredge<-dredge(NB2fit,rank = "BIC")
head(NB2dredge)
NB2best<-get.models(NB2dredge,subset=1)[[1]]
#summary(NB2best)
```
Ranked by BIC, the best model includes only the area interaction (weight: .99).

Assignment 2:

1.
```{r}
obsdat$BUM=NULL
newdata<-expand.grid(Year=factor(levels(obsdat$Year)),
                    area=factor(levels(obsdat$area)),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(NB2best,
              newdata = newdata,
              se.fit = TRUE,
              type="response")
newdata$fit<-temp$fit
newdata$se<-temp$se.fit
ggplot(newdata,aes(x=as.numeric(as.character(Year)),y=fit,ymin=fit-se,ymax=fit+se,color=area))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)",
       main="Negative binomial fit")
  
```
Something goes wrong here. I tried removing the BUM column, rearranging the expand.grid arguments, can't figure it out.

Tweedie

```{r}

TweedieFit<-glmmTMB(SWO/sets~(Year+area+season)^2,
                  data=obsdat,
                  family="tweedie",
                na.action = "na.fail")
#Must set na.action to "na.fail"

TweedieDredge<-dredge(TweedieFit,rank = "BIC")
head(TweedieDredge)
TweedieBest<-get.models(TweedieDredge,subset=1)[[1]]
summary(TweedieBest)
plot(simulateResiduals(TweedieBest))
```

Residual plot is slightly different, but I think looks ok.

2.
```{r}
newdata2<-expand.grid(Year=factor(levels(obsdat$Year)),
                    area=factor("N"),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(TweedieBest,
              newdata = newdata2,
              se.fit = TRUE,
              type="response")
newdata2$fit<-temp$fit
newdata2$se<-temp$se.fit
bind_rows(list(NB2=filter(newdata,area=="N"),
          Tweedie=newdata2),.id="type") %>%
ggplot(aes(x=as.numeric(as.character(Year)),y=fit,ymin=fit-se,
           ymax=fit+se,color=type))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)")

```

definitely not right

3.
```{r}
lnorm.mean=function(x1,x1e) {
  exp(x1+0.5*x1e^2)
}

lnorm.se=function(x1,x1e) {
  ((exp(x1e^2)-1)*exp(2*x1+x1e^2))^0.5
}
lo.se=function(x1,x1e,x2,x2e) {
  if(length(x1[!is.na(x1) &!is.na(x2)])>1 & length(unique(x1))>1 & length(unique(x2))>1 )
   cor12=cor(x1[!is.na(x1) &!is.na(x2)],x2[!is.na(x1) &!is.na(x2)]) else cor12=0
  (x1e^2 * x2^2 +x2e^2*x1^2+2*x1*x2*cor12*x1e*x2e)^0.5
}

```

```{r}
obsdat<-read.csv("BUM2.csv") %>%
 mutate(Year=factor(Year),
        season=factor(season))
ggplot(obsdat,
       aes(x=Year,y=BUM))+
  stat_summary(fun.data="mean_se")
ggplot(obsdat,aes(x=BUM))+
  geom_histogram(binwidth=1)
```

Part 1. binomial


```{r}
obsdat$present<-ifelse(obsdat$BUM>0,1,0)
BinomFit<-glmmTMB(present~Year*season,
                  data=obsdat,
                  family="binomial",
                na.action = "na.fail")
BinDredge<-dredge(BinomFit,rank = "BIC")
head(BinDredge)
BinBest<-get.models(BinDredge,subset=1)[[1]]
summary(BinBest)
plot(simulateResiduals(BinBest))

```

Part 2. log(CPUE) positive only

```{r}
posdat<-filter(obsdat,present==1)%>%
  mutate(cpue=BUM/sets)
cpueFit<-glmmTMB(log(cpue)~Year*season,
                  data=posdat,
                  family="gaussian",
                na.action = "na.fail")
cpueDredge<-dredge(cpueFit,rank = "BIC")
head(cpueDredge)
cpueBest<-get.models(cpueDredge,subset=1)[[1]]
summary(cpueBest)
plot(simulateResiduals(cpueBest))

```

Calculate index

```{r}
newdata3<-expand.grid(Year=factor(levels(obsdat$Year)),
                    season=factor("1"),
                    sets=1)  #1 set to get CPUE
temp<-predict(BinBest,
              newdata = newdata3,
              se.fit = TRUE,
              type="response")
newdata3$binomial<-temp$fit
newdata3$binomial.se<-temp$se.fit
temp<-predict(cpueBest,
              newdata = newdata3,
              se.fit = TRUE) #response is on log scale
#lognormal conversion functions to get mean if positive
newdata3$pos.cpue<-lnorm.mean(temp$fit,temp$se.fit)
newdata3$pos.cpue.se<-lnorm.se(temp$fit,temp$se.fit)
#index
newdata3<-newdata3 %>%
  mutate(fit=binomial*pos.cpue,
         se=lo.se(binomial,binomial.se,pos.cpue,pos.cpue.se))
```

Compare fits

```{r}
bind_rows(list(NB2=filter(newdata,area=="N"),
          Tweedie=newdata2,
          DeltaLN=newdata3),.id="type") %>%
ggplot(aes(x=as.numeric(as.character(Year)),y=fit,ymin=fit-se,
           ymax=fit+se,color=type))+
  geom_point(alpha=0.8)+
  geom_errorbar(width=0.1,alpha=0.8)+
  labs(x="Year",
    y="Standardized CPUE (±SE)")
```

The trend over time, based on the delta-lognormal of the BUM2 data at least, shows a decrease in CPUE.