---
title: "CPUE project 1"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r,message=FALSE}
library(tidyverse)
library(glmmTMB)
library(MuMIn)
library(DHARMa)
theme_set(theme_bw())
```

For this question, we will use the obsdat dataset, but with the swordfish column (SWO) rather than BUM.  

1.Fit a Poisson model with your choice of predictor variables, along with the sets offset, and look at the DHARMa residuals. Does it fit well?

```{r}
obsdat<-read_csv("obsdatSwo.csv")%>%
  mutate(Year=factor(Year),
         season=factor(season),
         area=factor(area))
Poisson1<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="poisson")
plot(simulateResiduals(Poisson1))
```

Looks very bad.

2. Fit a negative binomial 2 model  and look at the DHARMa residuals. Does it fit well?  


```{r}
NB2<-glmmTMB(SWO~Year+area+season+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2")
plot(simulateResiduals(NB2))
```
Looks much better. 

3. Use the dredge function to find the BIC best model starting the model with all two way interactions. Which model is best?


```{r}
NB2<-glmmTMB(SWO~(Year+area+season)^2+offset(log(sets)),
                  data=obsdat,
                  family="nbinom2",
             na.action="na.fail")
NB2dredge<-dredge(NB2,rank="BIC")
NB2dredge
```

Best model has area only. 


