---
title: "Likelihood for linear regression"
format: html
editor: visual
---

## Fitting models to data with Sums of Squares

Linear regression works with the method of **least squares**, which finds the values of the slope (a) and intercept (b) that minimize the difference between measured and predicted y. Here is some example data

```{r,message=FALSE}
library(tidyverse)
ggplot(mpg,aes(x=hwy,y=cty))+
  geom_point()+
  stat_smooth(method="lm")+
  labs(x="Highway miles per gallon",y="City miles per gallon",
       title="Fuel efficiency regression")
```

1.  The regression equation is that the predicted value of *y*, $\hat y=a+bx$

    The sums of squares algorithm is:

    1.  Guess at the value of a, and b.
    2.  Calculate $\hat y=a+bx$
    3.  Calculate the sum over all the data points of squared differences between y and $\hat y$, $SS=\sum{(y-\hat y)^2}$
    4.  Use a numerical optimizer function like *optim* to find the values of *a* and *b* that minimize SS.
    5.  This is the best least squares estimate, which is the best fit linear regresion.

Here it is in an R function.

```{r}
RegressionSS<-function(par,x,y) {
  a<-par["a"]
  b<-par["b"]
  y_hat<-a+b*x
  SS<-sum((y-y_hat)^2)
  return(SS)
}
fit<-optim(par=c(a=0,b=1),
           fn=RegressionSS,
           x=mpg$hwy,
           y=mpg$cty)
fit
```

This gave us the same answer we would get from using the lm function in R.

```{r}
lm1<-lm(cty~hwy,data=mpg)
summary(lm1)
```

## Fitting models to data with Likelihood

Likelihood of the parameters is the probability of getting the y data, give the x data and parameter values.

$L(a,b,\sigma^2)=P(y_i|a,b,x_i)\sim normal(\hat y_i, \sigma^2)$

The formula for the normal distribution is:

$P(y|\mu,\sigma^2)=\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(y-\mu)^2}{2\sigma^2}}$

Because data points are independent, the likelihood multiplies across the data points.

$L(\mu,\sigma^2|y) = \prod_{i=1}^{n}P(y_i|\mu,\sigma^2) = \prod_{i=1}^{n}\frac{1}{\sqrt{2\pi\sigma^2}}e^{-\frac{(y_i-\mu)^2}{2\sigma^2}}$

Taking the log, we get.

$\log L(\mu,\sigma^2|y) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^{n}(y_i-\mu)^2$

So, maximizing log likelihood is the same as minimizing sums of squares, plus some scaling parameters. In other words, the best fit parameters have the highest likelihood of having generated the data.

So, we can fit the model with likelihood as well. The likelihood algorithm is:

1.  Make a function that calculates the negative log likelihood.
2.  Minimize the function

Note that we have one more parameter than when we used SS, the residual variance, which is how spread out the data are around the regression line. We need this to calculate the normal probabilities.

```{r}
RegressionLL<-function(par,x,y) {
  a<-par["a"]
  b<-par["b"]
  y_hat<-a+b*x
  LL<-sum(dnorm(y,mean=y_hat,sd=1,log=TRUE))
  return(-LL)
}
fit<-optim(par=c(a=0,b=1),
           fn=RegressionLL,
           x=mpg$hwy,
           y=mpg$cty)
fit
```

This is assuming that the y data are normally distributed around the mean predicted by $\hat y$ with variance $\sigma^2$.

```{r}
# Demonstration plot:
# y_i ~ Normal(mu_i, sigma), where mu_i = b0 + b1*x_i
b0 <- 2
b1 <- 0.8
sigma <- 1.1

# Line domain
x_grid <- tibble(x = seq(0, 10, length.out = 300)) %>%
  mutate(mu = b0 + b1 * x)

# Points along the line where we'll draw vertical normal densities
x_marks <- tibble(x = c(1.5, 3.5, 5.5, 7.5, 9.0)) %>%
  mutate(mu = b0 + b1 * x)

# Build vertical density "ribbons" centered at each mu_i
# We parameterize over y, compute dnorm(y | mu_i, sigma), then convert to horizontal width.
density_curves <- x_marks %>%
  rowwise() %>%
  do({
    xi <- .$x
    mui <- .$mu
    y_vals <- seq(mui - 3 * sigma, mui + 3 * sigma, length.out = 200)
    tibble(
      x0 = xi,
      mu = mui,
      y = y_vals,
      dens = dnorm(y_vals, mean = mui, sd = sigma)
    )
  }) %>%
  ungroup() %>%
  mutate(
    # Scale density to visible horizontal width in x units
    x = x0 + dens * 0.9
  )

# Optional: one simulated observation at each x_mark
set.seed(713)
obs <- x_marks %>%
  mutate(y = rnorm(n(), mean = mu, sd = sigma))

ggplot() +
  # regression line E[y|x] = b0 + b1*x
  geom_line(data = x_grid, aes(x, mu), linewidth = 1.1, color = "black") +
  
  # vertical normal density shapes at selected x values
  geom_path(
    data = density_curves,
    aes(x = x, y = y, group = x0),
    color = "#2C7FB8",
    linewidth = 0.9
  ) +
  
  # center lines showing each conditional mean mu_i
  geom_segment(
    data = x_marks,
    aes(x = x, xend = x, y = mu - 3 * sigma, yend = mu + 3 * sigma),
    linetype = "dashed",
    color = "gray55",
    linewidth = 0.4
  ) +
  
  # optional simulated points
  geom_point(data = obs, aes(x, y), size = 2, color = "#D95F0E") +
  
  labs(
    x = "x",
    y = "y",
    title = "Linear regression assumption"
  ) +
  theme_minimal(base_size = 12)
```

Likelihood is much more general than sums of squares, and can be used with distributions other than the normal.

## In class examples

1.  Using the human data from the exponential growth module, fit a linear regression with likelihood
2.  Now fit the non-linear version. In other words, don't log transform. Use the model directly, assuming the data are normally distributed around the un-transformed exponential curve.
