---
title: "Markdown and Quarto tutorial"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
---

## R Markdown and Quarto

Markdown and Quarto are code formats that automatically generate documents 
from R, using Latex to make formatted text, tables, equations, tables of
contents, references, etc. You can make include your R code if you are
making a demonstration of your methods, or hide it if you are making a 
report. 

Quarto is the newer format which as more features such as easier biblographies ,
so you should use it, but Markdown is still widely used. The basic syntax is 
identical. See the RStudio cheat sheets
in the help menu for more information on Markdown, as well as the online documentation for
[Quarto](https://quarto.org/docs/tools/rstudio.html) and
[Markdown](https://rmarkdown.rstudio.com/)

This is a 10 minute video on getting started [Zoom recording](https://miami.zoom.us/rec/share/66N3YPZgCJspmKmaXsa3szOPhHz9CW9JbNSGRYoSDlJ2YiAKOoIShOz2Rc8cslGe.R5W2gM20JjQC4se_?startTime=1788790749000
Passcode: pbliKp&3)

To start a Markdown or Quarto document, go to File->New File->R Markdown 
or Quarto Document. You can choose a title, author, and output format when 
make the file, and you can change it later. This new file is not saved, so
you must save it in your desired working directory. Note that you can't 
easily change working directories when using Quarto or Markdown, so make 
sure your file and the data you need are in the same directory. 

You may run the code interactively in the document. When you are done writing
or debugging, hit the the knit button (Markdown) or the render button (Quarto)
in the top center of the screen to generate the output document. The output
file will be saved in your working directory.You may also see the output in
your Viewer window (lower right panel) in RStudio if go to Tools-->Global
Options-->R Markdown and check "Show outview preview in Viewer Pane".



## The yaml heading

At the top of the .Rmd or .qmd file, there is a heading with the title, author, date, 
and output format. Here is a sample yaml heading for Markdown:

````text
---
title: "Markdown and Quarto tutorial"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---
````

and here is a sample yaml heading for Quarto:

````text
---
title: "Quarto example file"
format: html
editor: visual
---
````
If you want to print to pdf rather than html, you need a 
Latex editor. If you don't have one, install the R library 
[tinytex](https://yihui.org/tinytex/).  

The yaml heading can also include information about file formatting, such as whether
to make a table of contents, and whether to include a bibliography.

In either Markdown or Quarto, RStudio lets you chose the `visual` or the 
`source` editor (upper left of your screen), where `source` is the plain code
and `visual` lets you see at least some of the formatting. 

## Code chunks. 

Your R code must be enclosed in code chunks, which are indicated by 
three backticks followed by {r} and then the code. For example:

````text
```{r}
# Your R code here
```
````

Everything between the backticks is treated as R code. You can run the chunks
one at a time by clicking the green triangle in the upper right of the
chunk. You can also run all chunks at once by clicking "Run" and then 
"Run All Chunks". The output will show  up under the chunk in your document, 
so you no longer need to look at the console or plots window.

For each chunk, you can set options to control the following (these all default 
to TRUE):

- `echo=FALSE` hides the code but keeps the output
- `message=FALSE` suppresses messages
- `warning=FALSE` suppresses warnings
- `eval=FALSE` shows code without running it

For example:

````text
```{r,message=FALSE,warning=FALSE}
library(tidyverse)
```
````
Would load the tidyverse library without showing all the messages about 
conflicting functions that are printed when you load the library. 

## Formatting the output


You can write regular text, and use markdown formatting to make headings, bold,
italics, lists, etc. 

- `#` creates a first-level heading
- `##` creates a second-level heading
- `###` creates a third-level heading
- `*italic*` or `_italic_`
- `**bold**`
- `-` or `*` for bullet lists
- `1.` for numbered lists

## Embedding formulas

You can also use latex code to embed formulas, using two dollar signs for centered 
formulas, and on dollar sign for inline formulas.  For example

````text
$$
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2}
$$ 

where $\bar{x}$ is the mean of the $n$ values $x_i$.
````
renders as 
$$
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i-\bar{x})^2}
$$ 

where $\bar{x}$ is the mean of the $n$ values $x_i$.


## Embedding R code in text output

You can also embed R code in your output. This is useful for 
putting the results of an analysis in the text with the right number of significant 
figures. For example, if I did the following analysis, I might want to sum it up in 
words at the end. 

```{r,message=FALSE}
library(tidyverse)
ggplot(cars, aes(x=speed,y=dist)) + 
  geom_point() + 
  geom_smooth(method="lm")
carslm<- lm(dist~speed,data=cars)
```


This text: 

```{r}
#  The slope of the regression line is `r round(coef(carslm)[2],2)` and the intercept is `r round(coef(carslm)[1],2)`.
```

renders as:

The slope of the regression line is `r round(coef(carslm)[2],2)` and the
intercept is `r round(coef(carslm)[1],2)`.

## Tables 

You can make simple tables directly in the text by using this format:

````text

| Group    | Mean | SD   |
|----------|------|------|
| Control  | 36.0 | 7.4  |
| Hormone  | 45.4 | 10.8 |

````

which renders as:

| Group    | Mean | SD   |
|----------|------|------|
| Control  | 36.0 | 7.4  |
| Hormone  | 45.4 | 10.8 |


You can also use the kable or kableExtra libraries to make nicely formatted
tables from your R output.  For example, to have the right number
of significant digits:

```{r,message=FALSE}
library(kableExtra)
mtcars |> 
  group_by(cyl) |>
  summarize(mean_mpg=mean(mpg), sd_mpg=sd(mpg)) |> 
  kable(digits=2, caption="Table 1. Summary of mpg by number of cylinders") |>
  kable_styling(bootstrap_options = c("striped"))
```

## Conclusion

I recommend you use Quarto (preferred) or Markdown for homework,
as it is easier to keep track of your notes and code in one place,
with the output visible. You can 
copy and paste from the knitted/rendered html file directly into Canvas, and 
you will have both the code and the output in your answer. 

I am posting my Markdown and Quarto code for all the lectures, so you can
see some example syntax.


