---
title: "Surveys"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r,message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

1. Using the data from the shark BLL survey, calculate the mean abundance in each year SPHYRNA MOKARRAN (great hammerhead), assuming a simple random sample.

```{r}
station<-read.csv("Station_Data.csv") %>%
  mutate(date=as.Date(DATE),
         year=year(date))%>%
  filter(year>2000)

catch<-read.csv("Catch_Data.csv")

```

Make a new column for species and count catch by species

```{r}
catchSum<-catch %>%
  mutate(Species=paste(GENUS,SPECIES)) %>%
  filter(Species %in% c("RHIZOPRIONODON TERRAENOVAE","SPHYRNA MOKARRAN"))%>%
  group_by(SEAMAPSTATION,CID,Species) %>%
  summarize(count=n())%>%
  pivot_wider(names_from=Species,values_from=count,values_fill = 0)
StationData<-left_join(station,catchSum,by=c("SEAMAPSTATION","CID"))
StationData<-StationData %>% 
  mutate(greatHammer=replace_na(`SPHYRNA MOKARRAN`,0),
         AtlSharpnose=replace_na(`RHIZOPRIONODON TERRAENOVAE`,0))
```

Get annual mean numbers

```{r}
ggplot(StationData,aes(x=factor(year),y=AtlSharpnose))+
  geom_boxplot()
yearSum<-StationData %>% 
  group_by(year) %>%
  summarize(meanAS=mean(greatHammer),
            seAS=sd(greatHammer)/sqrt(n()))  
ggplot(yearSum,aes(x=year,y=meanAS,ymin=meanAS-2*seAS,ymax=meanAS+2*seAS))+
  geom_point()+
  geom_errorbar(width=0.1)+
  ylab("Mean count of Great hammerhead")
```



2. Using the impala data with distance sampling, use AIC to find the best functional form to model detection probability as a function of distance. Plot the detection functions. Does the AIC-best also look like the best fit?  What is the total number in the sampled area and the density in the best model? 


```{r}
library(Distance)
impala<-read.csv("impala.csv") %>%
  rename(distance=x)%>%  #column must be called distance (in meters)
  mutate(Effort=60000,
         Region.Label="combined",
         Sample.Label="combined")  #total length of transects in meters
ggplot(impala,aes(x=distance))+geom_histogram()

```

Both distance and total effort are in meters. We don't have information on individual transects or habitat area in the region, so we can only calculate density. 

```{r}
impalaHN<-ds(impala,
             key="hn",
             adjustment = NULL)
cutpoints=seq(0,450,25)
plot(impalaHN, breaks=cutpoints, main="Half-normal")
impalaUC<-ds(impala,
             key="unif",
             adjustment = "cos")
plot(impalaUC, breaks=cutpoints, main="Uniform cosine")
impalaHP<-ds(impala,
             key="hr",
             adjustment = "poly")
plot(impalaHP, breaks=cutpoints, main="Hazard polynomial")

AIC(impalaHN,impalaUC,impalaHP)
```

uniform cosine is best

```{r}
summary(impalaUC)
```


