---
title: "Surveys"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r,message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

### Simulated data for stratification

```{r}
set.seed(10)
meanVals<-c(1,2.4,3.5,10)
area<-expand.grid(x=1:100,y=1:100) %>%
  mutate(strata=case_when(x<=50&y<=50~1,
                          x<=50&y>50~2,
                          x>50&y<=50~3,
                          x>50&y>50~4),
         N=rpois(10000,meanVals[strata]))  #random distribution
ggplot(area,aes(x=x,y=y,fill=N))+
  geom_tile()+
  annotate("text",x=c(25,25,75,75),y=c(25,75,25,75),label=1:4,color="red")
```

Real population

```{r}
sum(area$N)
```



Make a simple random sample


```{r}
n<-50
dataSimple<-sample(1:nrow(area),size=n,replace=FALSE)
ggplot(area,aes(x=x,y=y,fill=N))+
  geom_tile()+
  geom_point(data=area[dataSimple,],aes(x=x,y=y),color="red")

simple<-area[dataSimple,] %>% 
  summarize(meanCount=mean(N),
            seCount=sd(N)/sqrt(n()),
            n=n(),
            N=meanCount*nrow(area),
            seN=seCount*nrow(area),
            lci=N-2*seN,
            uci=N+2*seN)
simple %>% mutate(across(where(is.numeric),round,2))

```

Stratified with same sample size

```{r}
SamplesPerStrata<-round(n*meanVals/sum(meanVals))  #variance=mean for Poisson
SamplesPerStrata
strataSample<-NULL
for(i in 1:4) strataSample<-
  c(strataSample,sample(which(area$strata==i),
                        size=SamplesPerStrata[i],
                        replace=FALSE))
table(area$strata[strataSample])
ggplot(area,aes(x=x,y=y,fill=N))+
  geom_tile()+
  geom_point(data=area[strataSample,],aes(x=x,y=y),color="red")

stratified<- area[strataSample,] %>% 
  group_by(strata)%>%
  summarize(meanCount=mean(N),
            seCount=sd(N)/sqrt(n()),
            n=n(),
            area=nrow(area)/4,
            N=meanCount*area,
            seN=seCount*area) %>%
  ungroup()

stratified %>% mutate(across(where(is.numeric),round,2))
stratTotal<-stratified %>% summarize(N=sum(N),
                        seN=sqrt(sum(seN^2)),
                        lci=N-2*seN,
                        uci=N+2*seN)
stratTotal %>% mutate(across(where(is.numeric),round,2))

bind_rows(list(simple=simple,
               stratified=stratTotal),
          .id="Method")%>%
  ggplot(aes(x=Method,y=N,ymin=lci,ymax=uci))+
  geom_point()+
  geom_errorbar(width=0.1)+
  geom_hline(yintercept=sum(area$N),color="red")
```
With finite sample corrections. 

```{r}
simpleFinite<-area[dataSimple,] %>% 
  summarize(meanCount=mean(N),
            seCount=sd(N)/sqrt(n()),
            f=n/nrow(area),
            n=n(),
            N=meanCount*nrow(area),
            seN=(1-f)*seCount*nrow(area),
            lci=N-2*seN,
            uci=N+2*seN)
simpleFinite %>% mutate(across(where(is.numeric),round,2))
```

Makes no difference because sample is a small part of the whole population

### Simple random sample and stratified random

Grand Banks Yellowtail Flounder in 2006 from Brad Stevens

```{r}
tows <- read.csv("GB_YTF_2006_Tows2.csv")
head(tows)
strata<-read.csv("strata.csv")
strata
ggplot(tows,aes(x=factor(Stratum),y=Num,color=Season))+
  stat_summary(position=position_dodge(width=0.2))
with(tows,table(Season,Stratum))
```

Simple random sample in stratum 1160, spring only

```{r}
strataSum<-tows %>% 
  filter(Season=="Spring",Stratum==1160)%>%
  group_by(Stratum) %>%
  summarize(meanCount=mean(Num),
            seCount=sd(Num)/sqrt(n()),
            n=n(),
            meanDensity=mean(Num/area),
            seDensity=sd(Num/area)/sqrt(n()),
            sampled.area=sum(area))
#Or, as tows are all the same area (0.01)
strataSum<-strataSum %>%
  mutate(meanDensity2=meanCount/0.01,
         seDensity2=seCount/0.01)
# Use total area to get total numbers
strataSum<-strataSum %>% 
  mutate(N=meanDensity*2980,
         seN=seDensity*2980)
```


Total average density and total numbers across seasons and spatial stratas. 
Calculate totals per stratum, use left_join to add total areas to each stratum


```{r}
stratSeasonSum<-tows %>% group_by(Stratum,Season) %>%
  summarize(meanCount=mean(Num),
            seCount=sd(Num)/sqrt(n()),
            n=n(),
            meanDensity=mean(Num/area),
            seDensity=sd(Num/area)/sqrt(n()),
            sampled.area=sum(area)) %>%
  left_join(strata,by="Stratum") %>%
  mutate(N=meanDensity*area,
         seN=seDensity*area) %>%
  ungroup()
view(stratSeasonSum)
```

Some strata are understudied, so can't be estimated. For now, fill in 0 for the variances to get the sum in each season

```{r}
TotalSeason<-stratSeasonSum %>%
  mutate(seN=replace_na(seN,0),
         seDensity=replace_na(seDensity,0)) %>%
  group_by(Season)%>%
  summarize(N=sum(N),
            seN=sqrt(sum(seN^2)),
            lciN=N-2*seN,
            uciN=N+2*seN,
            meanDensity=sum(meanDensity*area)/sum(area),
            seDensity=sqrt((sum(seDensity^2*area)/sum(area))))

TotalSeason
```

For comparison, unstratified sample in each season (not really correct)

```{r}
simpleSum<-tows %>%
  group_by(Season)%>%
  summarise(meanCount=mean(Num),
            seCount=sd(Num)/sqrt(n())) %>%
  mutate(N=meanCount*sum(strata$area)/0.01,
         seN=seCount*sum(strata$area)/0.01,
         lciN=N-2*seN,
         uciN=N+2*seN)
simpleSum

```

Comparing estimates

```{r}
bind_rows(list(simple=simpleSum,
               stratified=TotalSeason),.id="Method") %>%
  ggplot(aes(x=Season,y=N,ymin=lciN,ymax=uciN,color=Method,group=Method))+
  geom_errorbar(position=position_dodge(width=0.1),width=0.1)+
  geom_point(position=position_dodge(width=0.1))
```

In this case it doesn't make much difference if you use simple or stratified



### Example 2. Shark bottom longline data

https://www.fisheries.noaa.gov/inport/item/29395
https://catalog.data.gov/dataset/shark-and-red-snapper-bottom-longline-survey

 Strata: 9 – 55 m, 55 – 183 m, and 183 – 366 m, but we will ignore this for now

```{r}
station<-read.csv("Station_Data.csv") %>%
  mutate(date=as.Date(DATE),
         year=year(date))%>%
  filter(year>2000)

catch<-read.csv("Catch_Data.csv")
head(catch)
view(station)
 
```

Make a new column for species and count catch by species

```{r}
catchSum<-catch %>%
  mutate(Species=paste(GENUS,SPECIES)) %>%
  filter(Species %in% c("RHIZOPRIONODON TERRAENOVAE","LUTJANUS CAMPECHANUS"))%>%
  group_by(SEAMAPSTATION,CID,Species) %>%
  summarize(count=n())%>%
  pivot_wider(names_from=Species,values_from=count,values_fill = 0)
view(head(catchSum))
StationData<-left_join(station,catchSum,by=c("SEAMAPSTATION","CID"))
StationData<-StationData %>% 
  mutate(redSnapper=replace_na(`LUTJANUS CAMPECHANUS`,0),
         AtlSharpnose=replace_na(`RHIZOPRIONODON TERRAENOVAE`,0))
view(StationData)
```

Get annual mean numbers

```{r}
ggplot(StationData,aes(x=factor(year),y=AtlSharpnose))+
  geom_boxplot()
yearSum<-StationData %>% 
  group_by(year) %>%
  summarize(meanAS=mean(AtlSharpnose),
            seAS=sd(AtlSharpnose)/sqrt(n()))  
ggplot(yearSum,aes(x=year,y=meanAS,ymin=meanAS-2*seAS,ymax=meanAS+2*seAS))+
  geom_point()+
  geom_errorbar(width=0.1)+
  ylab("Mean count of Atlantic Sharpnose")
```


### Distance sampling

https://distancesampling.org/Distance/articles/lines-distill.html

https://distancesampling.org/Distance/articles/web-only/points/pointtransects-distill.html

```{r,message=FALSE}
library(Distance)
```

Line transect with wrens

```{r}
data(wren_lt)
ggplot(wren_lt,aes(x=distance))+
  geom_histogram(binwidth=5)+
  ggtitle("Wren distance from transect line")

ggplot(wren_lt,aes(x=as.numeric(Sample.Label)))+
  geom_bar()+
  labs(x="Transect number",
       y="Number of wrens observed",
       title="Counts of wrens by transect")
head(wren_lt)
```
Area is the total hectares in the study area. Effort is the length (km) of the transect. Distance is in m. Convert_units function gives units of distance, transect length and area. 

Best detection function found with ds()

```{r}
conversion.factor<-convert_units("meter", "kilometer", "hectare")
wren.hn <- ds(data=wren_lt, 
              key="hn", 
              adjustment=NULL, 
              convert_units=conversion.factor)
summary(wren.hn)
cutpoints <- c(0,5,10,15,20,30,40,50,65,80,100)
plot(wren.hn, breaks=cutpoints, main="Half normal model, wren line transects")

```
Try other functions

```{r}
wren.unif.cos <- ds(wren_lt, key="unif", adjustment="cos", convert_units=conversion.factor)
plot(wren.unif.cos, breaks=cutpoints, main="Uniform cosine model, wren line transects")
wren.hr.poly <- ds(wren_lt, key="hr", adjustment="poly", convert_units=conversion.factor)
plot(wren.hr.poly, breaks=cutpoints, main="Hazard polynomial, wren line transects")
AIC(wren.hn, wren.hr.poly, wren.unif.cos)

```

Check fit of best model

```{r}
gof_ds(wren.hr.poly)
```


Point transect with sparrows

```{r}

data(Savannah_sparrow_1980)
Savannah_sparrow_1980$Region.Label <- "Single_stratum"
ggplot(Savannah_sparrow_1980,aes(x=distance))+
  geom_histogram()
#?ds
```


Find appropriate detection function

```{r}
sasp.hn <- ds(data=Savannah_sparrow_1980, 
              key="hn", #half normal
              adjustment=NULL,#no adjustment based on fitting variables
              transect="point", #data is distance from point of observation. 
              #No sampling effort given
              convert_units=convert_units("meter", NULL, "hectare"),
              truncation="5%") #Leave out most distance 5%
summary(sasp.hn)
cutpoints <- c(0,5,10,15,20,30,40,max(Savannah_sparrow_1980$distance, na.rm=TRUE))
plot(sasp.hn, breaks=cutpoints, pdf=TRUE, main="Savannah sparrow point transect data.")
```
With a different detection function

```{r}
sasp.unif.cos <- ds(Savannah_sparrow_1980, key="unif", adjustment="cos",
                    transect="point",
                    convert_units=convert_units("meter", NULL, "hectare"), truncation="5%")

plot(sasp.unif.cos, breaks=cutpoints, pdf=TRUE, main="Savannah sparrow point transect data.")
```
Compare models

```{r}
AIC(sasp.hn, sasp.unif.cos)
```

Half normal is best

```{r}
gof_ds(sasp.hn)
```

Residuals look fine

