
```{r,message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

### Example 2. Shark bottom longline data

https://www.fisheries.noaa.gov/inport/item/29395
https://catalog.data.gov/dataset/shark-and-red-snapper-bottom-longline-survey

 Strata: 9 – 55 m, 55 – 183 m, and 183 – 366 m, but we will ignore this for now

```{r}
station<-read.csv("Station_Data.csv") %>%
  mutate(date=as.Date(DATE),
         year=year(date))%>%
  filter(year>2000)

catch<-read.csv("Catch_Data.csv")
head(catch)

 
```

Make a new column for species and count catch by species

```{r}
catchSum<-catch %>%
  mutate(Species=paste(GENUS,SPECIES)) %>%
  filter(Species %in% c("SPHYRNA MOKARRAN"))%>%
  group_by(SEAMAPSTATION,CID,Species) %>%
  dplyr::summarize(count=n())%>%
  pivot_wider(names_from=Species,values_from=count,values_fill = 0)
head(catchSum)
StationData<-left_join(station,catchSum,by=c("SEAMAPSTATION","CID"))
StationData<-StationData %>% 
  mutate(greatHammer=replace_na(`SPHYRNA MOKARRAN`,0))
```

Get annual mean numbers

```{r}
ggplot(StationData,aes(x=factor(year),y=greatHammer))+
  geom_boxplot()
yearSum<-StationData %>% 
  group_by(year) %>%
  dplyr::summarize(meanAS=mean(greatHammer),
            seAS=sd(greatHammer)/sqrt(n()))  
ggplot(yearSum,aes(x=year,y=meanAS,ymin=meanAS-2*seAS,ymax=meanAS+2*seAS))+
  geom_point()+
  geom_errorbar(width=0.1)+
  ylab("Mean count of Great Hammerheads")
```


### Distance sampling

https://distancesampling.org/Distance/articles/lines-distill.html

https://distancesampling.org/Distance/articles/web-only/points/pointtransects-distill.html

```{r,message=FALSE}
library(Distance)
```

Line transect with impalas

```{r}
impala<-read.csv("impala.csv") %>%
  dplyr::rename(distance=x)%>%  #column must be called distance (in meters)
  mutate(Effort=60000,
         Region.Label="combined",
         Sample.Label="combined")  #total length of transects in meters

data(impala)
ggplot(impala,aes(x=distance))+
  geom_histogram(binwidth=5)+
  ggtitle("impala distance from transect line")

ggplot(impala,aes(x=as.numeric(distance)))+
  geom_bar()+
  labs(x="Transect number",
       y="Number of impalas observed",
       title="Counts of impalas by transect")
head(impala)
```
Area is the total hectares in the study area. Effort is the length (km) of the transect. Distance is in m. Convert_units function gives units of distance, transect length and area. 

Best detection function found with ds()

```{r}
conversion.factor<-convert_units("meter", "kilometer", "hectare")
impala.hn <- ds(data=impala, 
              key="hn", 
              adjustment=NULL, 
              convert_units=conversion.factor)
summary(impala.hn)
cutpoints <- c(0,25,50,75,100,150,200,250,300,350)
plot(impala.hn, breaks=cutpoints, main="Half normal model, impala line transects")

```
Try other functions

```{r}
impala.unif.cos <- ds(impala, key="unif", adjustment="cos", convert_units=conversion.factor)
plot(impala.unif.cos, breaks=cutpoints, main="Uniform cosine model, impala line transects")
impala.hr.poly <- ds(impala, key="hr", adjustment="poly", convert_units=conversion.factor)
plot(impala.hr.poly, breaks=cutpoints, main="Hazard polynomial, impala line transects")
AIC(impala.hn, impala.hr.poly, impala.unif.cos)

```

The uniform cosine model fits best.

```{r}
gof_ds(impala.unif.cos)
```