---
title: "Pre-Module 3 Optional Tutorial: Exponents & Logarithms"
author: "Claude in GitHub Copilot, from prompts by Elizabeth A. Babcock"
format: html
editor: visual
---

> **Optional background preparation — estimated time: \~20 minutes**\
> This tutorial covers the mathematical tools you will need for Module 3. It focuses purely on exponents and logarithms as concepts; Module 3 itself will show you how they apply to population growth.

## Learning objectives

By the end of this tutorial you should be able to:

- Explain what an exponent is and apply the basic rules of exponents algebraically.
- Explain what a logarithm is and apply the basic rules of logarithms.
- Describe the mathematical constant *e* and why it arises from continuous growth.
- Use `exp()` and `log()` in R with confidence.

------------------------------------------------------------------------

## 1. What Is an Exponent?

An exponent tells you how many times to multiply a base number by itself:

$$b^n = \underbrace{b \times b \times \cdots \times b}_{n \text{ times}}$$

For example:

| Expression | Calculation | Result |
|------------|-------------|--------|
| $2^1$      | 2           | 2      |
| $2^2$      | 2 × 2       | 4      |
| $2^3$      | 2 × 2 × 2   | 8      |
| $2^{10}$   | 2 × 2 × …   | 1024   |

### Key rules of exponents

| Rule | Formula | Example |
|------------------------|------------------------|------------------------|
| Product rule | $b^m \times b^n = b^{m+n}$ | $2^3 \times 2^4 = 2^7 = 128$ |
| Quotient rule | $b^m / b^n = b^{m-n}$ | $2^5 / 2^2 = 2^3 = 8$ |
| Power of a power | $(b^m)^n = b^{m \times n}$ | $(2^3)^2 = 2^6 = 64$ |
| Zero exponent | $b^0 = 1$ | $5^0 = 1$ |
| Negative exponent | $b^{-n} = 1/b^n$ | $2^{-3} = 1/8 = 0.125$ |
| Fractional exponent | $b^{1/n} = \sqrt[n]{b}$ | $9^{0.5} = 3$ |

### Try it in R

```{r}
# Exponentiation in R uses ^
2^10          # 1024
2^(-3)        # 0.125
9^(0.5)       # 3  (square root)

# The product rule: b^m * b^n = b^(m+n)
2^3 * 2^4     # 128
2^(3 + 4)     # 128 — same answer
```

------------------------------------------------------------------------

## 2. What Is a Logarithm?

A logarithm is the **inverse** of exponentiation. It answers the question:

> *"To what power must I raise the base* $b$ to get $x$?"

$$\log_b(x) = n \quad \Longleftrightarrow \quad b^n = x$$

Examples:

- $\log_2(8) = 3$ because $2^3 = 8$
- $\log_{10}(1000) = 3$ because $10^3 = 1000$

### The natural logarithm

In ecology and population biology, we almost always use the **natural logarithm**, whose base is the mathematical constant $e \approx 2.718$. It is written $\ln(x)$ in mathematics and simply `log(x)` in R.

$$\ln(x) = \log_e(x)$$

::: callout-warning
**R pitfall.** `log()` in R is the *natural* logarithm (base $e$), **not** base 10. Use `log10()` for base-10 logs. In population dynamics, `log()` is almost always what you want.
:::

### Key rules of logarithms

| Rule | Formula | Why it matters |
|------------------------|------------------------|------------------------|
| Log of a product | $\ln(ab) = \ln(a) + \ln(b)$ | Turns multiplication into addition |
| Log of a quotient | $\ln(a/b) = \ln(a) - \ln(b)$ | Turns division into subtraction |
| Log of a power | $\ln(a^n) = n \cdot \ln(a)$ | Brings exponents down as multipliers |
| Log of 1 | $\ln(1) = 0$ |  |
| Log of $e$ | $\ln(e) = 1$ | Definition of the natural log |
| Inverse property | $\ln(e^x) = x$ and $e^{\ln(x)} = x$ | `log` and `exp` cancel each other |

### Try it in R

```{r}
log(1)            # 0
log(exp(1))       # 1  — log and exp cancel
exp(log(5))       # 5  — exp and log cancel
log10(1000)       # 3  — base-10 log

# Demonstrate the product rule
log(3 * 4)        # should equal log(3) + log(4)
log(3) + log(4)   # same answer

# Demonstrate the power rule
log(2^10)         # should equal 10 * log(2)
10 * log(2)       # same answer
```

------------------------------------------------------------------------

## 3. The Number *e*

$e \approx 2.71828$ is a fundamental mathematical constant. It appears whenever something grows or decays **continuously** at a constant proportional rate.

### Where does *e* come from?

Imagine a bank account that earns 100% interest per year. If interest is compounded more and more frequently, the final balance after one year approaches a limit:

| Compounding   | Formula                          | Result              |
|---------------|----------------------------------|---------------------|
| Once per year | $(1 + 1)^1$                      | 2.000               |
| Monthly       | $(1 + 1/12)^{12}$                | 2.613               |
| Daily         | $(1 + 1/365)^{365}$              | 2.7146              |
| Continuously  | $\lim_{n \to \infty}(1 + 1/n)^n$ | $e = 2.71828\ldots$ |

This limit defines $e$. Populations reproduce more-or-less continuously rather than all at once, so $e$ appears naturally in the equations you will encounter in Module 3.

```{r}
# e in R
exp(1)          # 2.718282

# Demonstrating the limit that defines e
n <- c(1, 12, 365, 1e6)
(1 + 1/n)^n     # approaches exp(1) as n increases
```

### *e* as a base for exponents

The function $e^x$ (written `exp(x)` in R) has a special property: its rate of change at any point equals its own value. This "self-referential" growth is exactly what happens when a population grows at a constant per-capita rate — which is why you'll see $e^{r_m t}$ in Module 3.

```{r}
# exp(x) in R
exp(0)    # 1
exp(1)    # e ≈ 2.718
exp(2)    # e^2 ≈ 7.389
exp(-1)   # 1/e ≈ 0.368  (decay)
```

------------------------------------------------------------------------

## 4. Quick Reference

| R function | What it does              | Example            |
|------------|---------------------------|--------------------|
| `exp(x)`   | Calculates $e^x$          | `exp(1)` → 2.718   |
| `log(x)`   | Natural log $\ln(x)$      | `log(2.718)` → 1.0 |
| `log10(x)` | Log base 10               | `log10(1000)` → 3  |
| `x^n`      | $x$ to the power $n$      | `2^8` → 256        |
| `sqrt(x)`  | Square root (= $x^{0.5}$) | `sqrt(9)` → 3      |

------------------------------------------------------------------------

## 5. Self-Check Questions

Try these before starting Module 3.

**Question 1.** Simplify $3^4 \times 3^{-2}$. What is the numerical answer?

<details>

<summary>Show answer</summary>

Using the product rule: $3^4 \times 3^{-2} = 3^{4+(-2)} = 3^2 = \mathbf{9}$.

In R: `3^4 * 3^(-2)` or `3^(4 + -2)`.

</details>

**Question 2.** What is $\ln(e^5)$? What rule did you use?

<details>

<summary>Show answer</summary>

By the inverse property, $\ln(e^5) = \mathbf{5}$. The log and exponent cancel because they are inverse operations. In R: `log(exp(5))`.

</details>

**Question 3.** Use the power rule to simplify $\ln(x^3)$.

<details>

<summary>Show answer</summary>

$\ln(x^3) = 3 \cdot \ln(x)$. This is the log-of-a-power rule: the exponent comes down as a multiplier.

</details>

**Question 4.** Without using R, what is $e^0$?

<details>

<summary>Show answer</summary>

$e^0 = \mathbf{1}$. Any number raised to the power zero equals 1 (the zero-exponent rule).

</details>

------------------------------------------------------------------------

## Summary

- **Exponents** describe repeated multiplication: $b^n$ multiplies $b$ by itself $n$ times.
- **Logarithms** are the inverse of exponents: $\ln(x)$ asks "to what power must I raise $e$ to get $x$?"
- The **natural logarithm** (base $e$) is the standard in ecology; in R it is `log()`.
- `log()` and `exp()` are inverses: they cancel each other.
- The constant $e \approx 2.718$ arises naturally from continuous compounding, making it the right base for continuous growth processes.

You are now ready for Module 3. Good luck!

------------------------------------------------------------------------

*MBE 713 Population Dynamics — Elizabeth A. Babcock*\
*Optional preparatory material for Module 3 (Exponential Growth)*
