---
title: "6. Leslie matrices"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r,message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

### Life table calculations

```{r,message=FALSE,warning=FALSE}
cohort<-read_csv("cohortLifeTable.csv")
ggplot(cohort,aes(x=x))+
  geom_line(aes(y=S_F,color="F"))+
  geom_line(aes(y=S_M,color="M"))+
  labs(x="Age",y="Survivors to beginning of age group")
ggplot(cohort,aes(x=x))+
  geom_line(aes(y=log(S_F),color="F"))+
  geom_line(aes(y=log(S_M),color="M"))+
  labs(x="Age",y="log(Survivors to beginning of age group")
```

Calculate number of deaths

```{r}
cohort$D_M<-cohort$S_M-c(cohort$S_M[2:10],0)
cohort$D_F<-cohort$S_F-c(cohort$S_F[2:10],0)
cohort
```

Calculate mortality rate

```{r,warning=FALSE}
cohort<-cohort |> mutate(
  q_M=D_M/S_M,
  q_F=D_F/S_F)
cohort
ggplot(cohort,aes(x=x)) +
  geom_line(aes(y=q_M,color="M"))+
  geom_line(aes(y=q_F,color="F"))+
  labs(x="Age",y="Mortality rate")
```

Vertical life table 


```{r}
vertical<-read_csv("vertical.csv")
vertical<-vertical |>
  mutate(Years_in_ageclass=x1-x0,
         annual_mortality_rate=Deaths_per_year_in_sample/Sample,
         Numbers_per_1000=NA)

vertical$Numbers_per_1000[1]<-1000
for(i in 2:nrow(vertical)) {
  vertical$Numbers_per_1000[i]<-
    vertical$Numbers_per_1000[i-1]*
    (1-vertical$annual_mortality_rate[i-1])^
     vertical$Years_in_ageclass[i-1]
}
#view(vertical)
ggplot(vertical,aes(x=x0,y=Numbers_per_1000))+
         geom_line()
write_csv(vertical,"verticalSolved.csv")
```

### Demographic models

Read in the squirrel data, lawn pre-intervention only

```{r}
squirrel<-read_csv("squirrel.csv")
squirrel<-squirrel |> mutate(lxmx=lx*mx)
squirrel
```

Calculate lifetime reproductive rate

```{r}
R0<-with(squirrel,sum(lxmx))
R0
```

Mean generation time

```{r}
Tc<-with(squirrel,sum(x*lxmx)/sum(lxmx))
Tc
```
Euler-Lotka equation

Births at time t are equal to the sum over age groups (x) number of females born x years ago, times their survival to age x (lx) times their fecundity at age x (mx).

$$
B_t= \sum_{x=0}^{x_{max}} B_{t-x} l_x m_x
$$

And population is growing (or declining) at rate $\lambda=e^{r_m}$. So births are 

$$
B_t=B_0\lambda^t
$$
Substituting


$$
B_0\lambda^t = \sum_{x=0}^{x_{max}} B_0\lambda^{t-x}l_x m_x
$$

$B_0 \lambda^t$ cancels out, leaving 

$$
1 = \sum_{x=0}^{x_{max}} \lambda^{-x} l_x m_x  
$$

Or equivalently

$$
1 = \sum_{x=0}^{x_{max}} e^{-r_m x} l_x m_x  
$$
Solve numerically for $r_m$

```{r}
EulerEquation<-function(rm,data) {
  (sum(exp(-rm*data$x)*data$lx*data$mx)-1)^2
}
rvals<-data.frame(r=seq(-0.1,0.2,by=0.01))
for(i in 1:length(rvals$r)) {
 rvals$Euler[i]<-EulerEquation(rvals$r[i],data=list(x=squirrel$x,
                        lx=squirrel$lx,
                        mx=squirrel$mx))
}
ggplot(rvals,aes(x=r,y=Euler))+geom_line()
```

Find value of rm for squirrel data

```{r}
solvedEuler<-optim(par=0.1,
              fn=EulerEquation,
              data=list(x=squirrel$x,
                        lx=squirrel$lx,
                        mx=squirrel$mx),
              method="L-BFGS-B")

solvedEuler
rm<-solvedEuler$par
rm
```


$r_m$ can also be approximated by just assuming each female produces $R_0$ offspring at age %T_c% so that 

$$r_c=log(R_0)/T_c$$
which is, for the squirrels

```{r}
rc=log(R0)/Tc
rc
```
This would be exactly correct for an annual species, and it is approximate for species that reproduce multiple times. 

Matrix version

```{r}
Pvals<-squirrel$lx[2:6]/squirrel$lx[1:5]
Fvals<-squirrel$mx[2:6]*Pvals
Leslie<-matrix(0,5,5)
Leslie[1,]<-Fvals
for(i in 1:4) Leslie[i+1,i]<-Pvals[i]
round(Leslie,3)
```

Calculate $r_m$ with matrix multiplication  %*%

```{r}
nyear<-100
N<-matrix(NA,5,nyear)
N[,1]<-rep(1,5)
for(i in 2:nyear) {
  N[,i]<-Leslie %*% N[,i-1]
}
N[1:5,1:5]
sumTab<-data.frame(t=1:nyear,N=colSums(N))
view(sumTab)
sumTab$lambda<-NA
sumTab$lambda[2:nyear]<-sumTab$N[2:nyear]/sumTab$N[1:(nyear-1)]

ggplot(sumTab,
       aes(x=t,y=N,group=1))+
  geom_line()

ggplot(sumTab,
       aes(x=t,y=lambda,group=1))+
  geom_line()

lambda<-sumTab$lambda[nyear]
lambda
rm<-log(lambda)
rm
```

The stable age structure can be calculated by dividing the numbers by the total in the last year.

```{r}
round(N[,nyear]/sumTab$N[nyear],3)
```

Eigenvalue method

```{r}
eigen(Leslie)
lambda2<-Re(eigen(Leslie)$values[1])  #Re for just real number
lambda2
#Stable age structure is first eigenvector divided by its sum
round(Re(eigen(Leslie)$vectors[,1]/
           sum(eigen(Leslie)$vectors[,1])),3)

```
### Elasticity analysis

```{r}
#Function to calculated left eigenvalue #https://stackoverflow.com/questions/14912279/how-to-obtain-right-eigenvectors-of-matrix-in-r
left_eigen<-function(A) {
  Re(eigen(t(Leslie))$vectors)
}
w<-Re(eigen(Leslie)$vectors)[,1]
v<-left_eigen(Leslie)[,1]
w/sum(w)
v/v[1]

elasticityFunc<-function(A) {
 w<-Re(eigen(A)$vectors)[,1]
 v<-left_eigen(A)[,1]
 w<-w/sum(w)
 v<-v/v[1]
 lambda<-Re(eigen(A)$values)[1]
 sensitivity<-outer(v,w)/sum(v*w)
 elasticity<-(A/lambda)*sensitivity
 elasticity
}
elasticityFunc(Leslie)

```
Calculate some summaries

Elasticity of reproductive output

```{r}
elasVals<-elasticityFunc(Leslie)

```


Show elasticity for reproduction compared to survival

```{r}
reproElasticity<-elasVals[1,]
survivalElasticity<-c(elasVals[2,1],elasVals[3,2],
                      elasVals[4,3],elasVals[5,4])
reproElasticity
survivalElasticity
elastSum<-data.frame(Age=rep(1:5,2),
                     Parameter=rep(c("Reproduction","Survival"),each=5),
                     Elasticity=c(reproElasticity,survivalElasticity,0))
ggplot(elastSum,aes(x=Age,y=Elasticity,fill=Parameter)) +
  geom_col(position = position_dodge())+
  scale_fill_manual(values=c("darkred","goldenrod"))
ggplot(elastSum,aes(x=Parameter,weight=Elasticity,fill=Parameter)) +
  geom_bar()+
  scale_fill_manual(values=c("darkred","goldenrod"))

```
Reproduction is more important as a whole, particularly or younger animals. 

Try changing a parameter to see how it influences lambda. 

```{r}
newLeslie<-Leslie
newLeslie[1,1]<-Leslie[1,1]*1.1
newLambda<-Re(eigen(newLeslie)$values[1])
(newLambda-lambda)/lambda
```

A 10% increase in F1 is a 3.4% increase in lambda

### R library popbio

https://www.r-bloggers.com/2013/04/demographic-analysis-using-the-popbio-library-and-some-other-fun-stuff/

Library does all this work for you.

```{r}
library(popbio)
eigen.analysis(Leslie)
```

### Monte Carlo estimation of uncertainty


```{r}
getLeslie<-function(lx,mx) {
  Pvals<-lx[2:6]/lx[1:5]
  Fvals<-mx[2:6]*Pvals
  Leslie<-matrix(0,5,5)
  Leslie[1,]<-Fvals
  for(i in 1:4) Leslie[i+1,i]<-Pvals[i]
  Leslie
}
squirrel.lx.cv<-0.02
squirrel.mx.cv<-0.05
nsim<-100000
lambdaSims<-NULL
for(sim in 1:nsim) {
  lx<-c(1,rnorm(5,squirrel$lx[2:6],squirrel.lx.cv*squirrel$lx[2:6]))
  mx<-rnorm(6,squirrel$mx,squirrel.mx.cv*squirrel$mx)
  A<-getLeslie(lx,mx)
  lambdaSims[sim]<-Re(eigen(A)$values[1])
}
summary(lambdaSims)
#or more efficiently
getLambda<-function() {
    lx<-c(1,rnorm(5,squirrel$lx[2:6],squirrel.lx.cv*squirrel$lx[2:6]))
    mx<-rnorm(6,squirrel$mx,squirrel.mx.cv*squirrel$mx)
    A<-getLeslie(lx,mx)
    Re(eigen(A)$values[1])
}
lambdaSims<-replicate(nsim,getLambda())
#summarize results
summary(lambdaSims)
mean(lambdaSims,na.rm=TRUE) 
confidenceInterval<-quantile(lambdaSims,c(0.025,0.975))
round(confidenceInterval,3)
ggplot(data.frame(lambda<-lambdaSims),aes(x=lambda,y=..density..))+
  geom_histogram(fill="darkgrey",color="black",bins=50)+
  geom_density(color="red")
```





