---
title: "6a. Introduction to Matrix Multiplication"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

```{r,message=FALSE}
library(tidyverse)
```

### Why do we need this?

In the Leslie matrix module (`6.LeslieMatrices.rmd`) we use matrices and matrix
multiplication to project a population forward in time, and to find the
population growth rate and stable age structure. If you have never worked
with matrices before, the notation and the operations can look confusing at
first. This short tutorial (about 20-30 minutes) will teach you just enough
matrix algebra to understand those examples. We will use simple, traditional
examples (not population models) to build the concepts from scratch. At the
end there are some practice questions you can use to check your understanding.

### What is a matrix?

A matrix is just a rectangular table of numbers, arranged in rows and
columns. For example, here is a matrix with 2 rows and 3 columns. We say this
matrix has dimensions 2 x 3 (rows x columns):

$$
A=\begin{bmatrix} 1 & 2 & 3 \\ 4 & 5 & 6 \end{bmatrix}
$$

In R, we can make this matrix like this:

```{r}
A<-matrix(c(1,2,3,4,5,6),nrow=2,byrow=TRUE)
A
```

Notice that `byrow=TRUE` tells R to fill in the numbers row by row (1,2,3
in the first row, 4,5,6 in the second row). If you leave that out, R fills in
column by column instead, which will give you a different matrix, so be
careful.

A matrix with only one column (or only one row) is often called a **vector**.
For example

$$
v=\begin{bmatrix} 10 \\ 20 \\ 30 \end{bmatrix}
$$
is a column vector with 3 rows and 1 column (dimensions 3 x 1).

```{r}
v<-matrix(c(10,20,30),nrow=3)
v
```

### Indexing a matrix

Each number in a matrix has a row number and a column number. We refer to the
number in row $i$ and column $j$ as $A_{i,j}$. So in matrix A above,
$A_{1,1}=1$, $A_{1,3}=3$, and $A_{2,2}=5$.

In R, we can pull out individual elements using square brackets, with the row
number first and the column number second.

```{r}
A[1,1]
A[1,3]
A[2,2]
```

You can also pull out a whole row or column by leaving the other index blank.

```{r}
A[1,]  #First row
A[,2]  #Second column
```

### Adding and subtracting matrices

You can add or subtract two matrices if (and only if) they have exactly the
same dimensions. You just add or subtract the numbers that are in the same
position.

$$
\begin{bmatrix} 1 & 2 \\ 3 & 4 \end{bmatrix} +
\begin{bmatrix} 10 & 20 \\ 30 & 40 \end{bmatrix} =
\begin{bmatrix} 11 & 22 \\ 33 & 44 \end{bmatrix}
$$

```{r}
B<-matrix(c(1,2,3,4),nrow=2,byrow=TRUE)
C<-matrix(c(10,20,30,40),nrow=2,byrow=TRUE)
B+C
```

### Multiplying a matrix by a single number

Multiplying a matrix by a single number (a "scalar") just multiplies every
element by that number.

```{r}
2*B
```

### Matrix multiplication: the key idea

Matrix multiplication (multiplying two matrices together) is NOT the same as
multiplying the matching elements together like we did with addition. This is
the part that is confusing the first time you see it, so let's go slowly.

To multiply matrix A (dimensions m x n) by matrix B (dimensions n x p), the
number of **columns** in A must equal the number of **rows** in B. The
result, called $A B$, will have dimensions m x p.

Each element of the result is found by taking a **row** of A and a
**column** of B, multiplying the matching entries together, and adding up
the products. This is sometimes called the "dot product" of the row and the
column.

#### Example: matrix times a vector

Let's start with the simplest useful case: a matrix times a column vector.
Suppose

$$
A=\begin{bmatrix} 2 & 0 & 1 \\ 1 & 3 & 2 \end{bmatrix}
\qquad
v=\begin{bmatrix} 5 \\ 10 \\ 1 \end{bmatrix}
$$

A has dimensions 2 x 3, and v has dimensions 3 x 1. Since A has 3 columns and
v has 3 rows, we can multiply them, and the result will be 2 x 1.

To get the first element of the result, take the first **row** of A,
(2, 0, 1), and the (only) **column** of v, (5, 10, 1). Multiply matching
elements and add:

$$
2\times5 + 0\times10 + 1\times1 = 10+0+1=11
$$

To get the second element, take the second row of A, (1, 3, 2), and again the
column of v, (5, 10, 1):

$$
1\times5 + 3\times10 + 2\times1 = 5+30+2=37
$$

So

$$
Av=\begin{bmatrix} 11 \\ 37 \end{bmatrix}
$$

Let's confirm this in R. Matrix multiplication in R uses the special symbol
`%*%`, NOT the ordinary `*`, because `*` multiplies matching elements
(element-by-element), while `%*%` does the row-times-column matrix
multiplication described above.

```{r}
A<-matrix(c(2,0,1,1,3,2),nrow=2,byrow=TRUE)
v<-matrix(c(5,10,1),nrow=3)
A
v
A %*% v
```

That matches our hand calculation.

#### Example: matrix times matrix

Now let's multiply two matrices together. Suppose

$$
A=\begin{bmatrix} 1 & 2 \\ 3 & 4 \end{bmatrix}
\qquad
B=\begin{bmatrix} 5 & 6 \\ 7 & 8 \end{bmatrix}
$$

Both are 2 x 2, so we can multiply them, and the answer will also be 2 x 2.
Each element of the answer comes from a row of A and a column of B:

- Row 1 of A, Column 1 of B: $1\times5+2\times7=5+14=19$
- Row 1 of A, Column 2 of B: $1\times6+2\times8=6+16=22$
- Row 2 of A, Column 1 of B: $3\times5+4\times7=15+28=43$
- Row 2 of A, Column 2 of B: $3\times6+4\times8=18+32=50$

So

$$
AB=\begin{bmatrix} 19 & 22 \\ 43 & 50 \end{bmatrix}
$$

```{r}
A<-matrix(c(1,2,3,4),nrow=2,byrow=TRUE)
B<-matrix(c(5,6,7,8),nrow=2,byrow=TRUE)
A %*% B
```

**Important:** matrix multiplication is usually NOT commutative, meaning
$AB$ is usually not the same as $BA$. Order matters!

```{r}
B %*% A
```

You can see this gives a different answer than `A %*% B`.

### A practical use: repeated multiplication to project forward in time

One very common use of matrix multiplication is to describe how a system
changes from one time step to the next, and then repeat the same
multiplication over and over to project many time steps into the future.

For example, suppose we have a very simple weather model where the weather
tomorrow depends only on the weather today. Let a vector $N_t$ describe the
"state" of the system (fraction of days that are Sunny or Rainy) at time $t$:

$$
N_t=\begin{bmatrix} Sunny_t \\ Rainy_t \end{bmatrix}
$$

And suppose there is a "transition matrix" $M$ that describes the
probability of moving from one state to another:

$$
M=\begin{bmatrix} 0.8 & 0.4 \\ 0.2 & 0.6 \end{bmatrix}
$$

Here, column 1 refers to today being Sunny, and column 2 refers to today
being Rainy. Row 1 is the probability that tomorrow is Sunny, and row 2 is
the probability that tomorrow is Rainy. So this matrix says: if today is
Sunny, there is an 80% chance tomorrow is Sunny and 20% chance tomorrow is
Rainy; if today is Rainy, there is a 40% chance tomorrow is Sunny and 60%
chance tomorrow is Rainy.

If today is Sunny for sure, $N_0=\begin{bmatrix}1\\0\end{bmatrix}$, we can
find tomorrow's state by matrix multiplication:

$$
N_1 = M N_0
$$

```{r}
M<-matrix(c(0.8,0.4,0.2,0.6),nrow=2,byrow=TRUE)
N0<-matrix(c(1,0),nrow=2)
N1<-M %*% N0
N1
```

To project two days ahead, we multiply again, using $N_1$ as the new
starting point:

$$
N_2 = M N_1 = M (M N_0)
$$

```{r}
N2<-M %*% N1
N2
```

We could keep doing this, one day at a time, to see how the probabilities
change over many days. Here we do it for 20 days using a loop, storing each
day's result in a matrix with one column per day.

```{r}
ndays<-20
N<-matrix(NA,nrow=2,ncol=ndays)
N[,1]<-c(1,0)  #Day 1: Sunny for sure
for(i in 2:ndays){
  N[,i]<-M %*% N[,i-1]
}
round(N,3)
```

Notice that after a while, the probabilities stop changing much from one day
to the next. This is called reaching a "stable" or "equilibrium" state. This
same idea -- repeatedly multiplying a matrix by a vector describing the
current state of a system to project forward in time, and watching the
system settle into a stable pattern -- is exactly what is done with the
Leslie matrix to project a population's age structure forward in time and
find its long-term growth rate and stable age distribution.

```{r}
weather<-data.frame(day=1:ndays,Sunny=N[1,],Rainy=N[2,])
ggplot(weather,aes(x=day))+
  geom_line(aes(y=Sunny,color="Sunny"))+
  geom_line(aes(y=Rainy,color="Rainy"))+
  labs(x="Day",y="Probability",color="")+
  theme_bw()
```

### Summary of key points

1. A matrix is a rectangular table of numbers with rows and columns.
2. You can add/subtract matrices of the same dimensions element by element.
3. Matrix multiplication ($A \times B$, written `A %*% B` in R) is NOT the
   same as multiplying matching elements. Instead, each element of the result
   comes from multiplying a **row** of A by a **column** of B and adding the
   products.
4. To multiply $AB$, the number of columns in A must equal the number of
   rows in B.
5. Matrix multiplication is generally not commutative: $AB \ne BA$.
6. Repeatedly multiplying a matrix by a vector (updating the vector each
   time) is a common way to project a system forward in time, and it often
   settles into a stable pattern after enough repetitions.

### Practice questions

Try these on paper first, then check your answers in R.

**Question 1.** What are the dimensions of the matrix

$$
Q=\begin{bmatrix} 3 & 1 & 0 \\ 2 & 5 & 4 \end{bmatrix}
$$
and what is the value of $Q_{2,3}$?

**Question 2.** Can you multiply a 3 x 2 matrix by a 2 x 4 matrix? If so,
what will be the dimensions of the result? Can you multiply a 3 x 2 matrix
by a 3 x 2 matrix (using `%*%`, not element-by-element multiplication)?

**Question 3.** By hand, calculate $Av$ where

$$
A=\begin{bmatrix} 1 & 2 \\ 0 & 1 \\ 3 & 1 \end{bmatrix}
\qquad
v=\begin{bmatrix} 2 \\ 5 \end{bmatrix}
$$

Then check your answer using R.

**Question 4.** By hand, calculate $AB$ where

$$
A=\begin{bmatrix} 2 & 1 \\ 0 & 3 \end{bmatrix}
\qquad
B=\begin{bmatrix} 1 & 4 \\ 2 & 0 \end{bmatrix}
$$

Then calculate $BA$. Are they the same? Check both in R.

**Question 5.** Explain in your own words why, if $N_t$ is a vector
describing the state of a system at time $t$, and $M$ is a transition
matrix, $N_{t+2} = M(MN_t)$ describes the state of the system two time steps
later. What is another way you could write $M(MN_t)$ using only one
multiplication symbol between $M$ and $M$?

```{r,eval=FALSE,echo=FALSE}
# Answers (do not show to students by default)
# Q1: Q is 2 x 3 (2 rows, 3 columns). Q[2,3] = 4
# Q2: Yes, a 3x2 times a 2x4 gives a 3x4 result. A 3x2 times a 3x2 with %*%
#     is NOT allowed (columns of first, 2, must equal rows of second, 3).
# Q3: Av = [1*2+2*5, 0*2+1*5, 3*2+1*5]' = [12,5,11]'
# Q4: AB = [2*1+1*2, 2*4+1*0; 0*1+3*2, 0*4+3*0] = [4,8;6,0]
#     BA = [1*2+4*0, 1*1+4*3; 2*2+0*0, 2*1+0*3] = [2,13;4,2]
#     Not the same.
# Q5: M(MN_t) applies the transition matrix once to get N_{t+1}=M N_t, then
#     applies it again to that result to get N_{t+2}. This can also be
#     written as M %*% M %*% N_t, i.e. (M %*% M) %*% N_t, using matrix M
#     squared, M^2 (though R does not use ^ for this; you would use
#     M %*% M).
```
