---
title: "4. Fisheries production models"
author: "Beth Babcock"
date: "`Sys.Date()`"
output: html_document
---


```{r,message=FALSE}
library(tidyverse)
theme_set(theme_bw())
```

### Fitting a model with Bayesian methods

Here is some data, to calculate a mean using Bayesian priors

```{r}
data<-data.frame(i=1:30,x=c(7.72, 4.79, 5.78, 4.89, 2.25, 4.17, 4.21, 4.88, 7.2, 6.53, 4.67, 4.49, 6.39, 6.11, 3.62, 3.59, 5.73, 6.54, 4.78, 6.76, 5.8, 
3.78, 5.68, 2.74, 7.87, 8.96, 4.27, 2.91, 6.14, 4.73))  
```

Now let's try some priors. First, an informative prior. Let's assume that we have good information about the value of mu from another study, so our prior has a mean of 2 and sd of 2.

```{r}
priormean<-2
priorsd<-2
posterior<-data.frame(mu=seq(0,10,0.01))
for(i in 1:nrow(posterior)) {
 posterior$LL[i]<-sum(dnorm(data$x,mean=posterior$mu[i],sd=2))
}
posterior$logprior<-dnorm(posterior$mu,mean=priormean,sd=priorsd)
posterior$logposterior<-posterior$LL*posterior$logprior
ggplot(posterior,aes(x=mu))+
  geom_line(aes(y=LL/max(LL),color="log-likelihood"))+
  geom_line(aes(y=logprior/max(logprior),color="log-prior"))+
  geom_line(aes(y=logposterior/max(logposterior),color="log-posterior"))+
  ggtitle("Posterior is prior times likelihood")

```
Same with a less informative prior

```{r}
priormean<-2
priorsd<-10
posterior<-data.frame(mu=seq(0,10,0.01))
for(i in 1:nrow(posterior)) {
 posterior$LL[i]<-sum(dnorm(data$x,mean=posterior$mu[i],sd=2))
}
posterior$logprior<-dnorm(posterior$mu,mean=priormean,sd=priorsd)
posterior$logposterior<-posterior$LL*posterior$logprior
ggplot(posterior,aes(x=mu))+
  geom_line(aes(y=LL/max(LL),color="log-likelihood"))+
  geom_line(aes(y=logprior/max(logprior),color="log-prior"))+
  geom_line(aes(y=logposterior/max(logposterior),color="log-posterior"))+
  ggtitle("Posterior is prior times likelihood")+
  labs(y="Probability density",x="Value of mu parameter",color="")
view(posterior)
```
Posterior is more similar to the likelihood

Now make a function to minimize to get the mode of the posterior distribution

```{r}
getNegLogPosterior<-function(theta,data) {
    LL<-dnorm(x=data$x, #data is an input, with a variable called x
              mean=theta["mu"], #mean is the parameter we want to calculate
              sd=2,
              log=TRUE) #fix standard deviation for this example
    logprior<-dnorm(theta["mu"],data$priormean,data$priorsd,log=TRUE)
    -sum(LL)-logprior  #return negative total log likelihood plus priors
}
getNegLogPosterior(c(mu=5.2),list(x=data$x,priormean=2,priorsd=2))
```

Use minimizer function optim to find the value of mu that has the lowest -logposterior

```{r}
modFit<-optim(par=c(mu=3),
              fn=getNegLogPosterior,
              data=list(x=data$x,priormean=2,priorsd=2),
              method="L-BFGS-B")
modFit$par

```


### Bayesian production models in R. 

Possible priors

```{r}
dlogunif<-function(x,minval,maxval,logTrans=TRUE) {
  if(minval==0) minval<-0.0000001
  prob<-1/(x*(log(maxval)-log(minval)))
  if(logTrans) return<-log(prob) else 
    return=prob
  return 
}
dlogunif(1,0,1000,logTrans=FALSE)

ggplot(data.frame(r=c(0,1)),aes(x=r))+
  geom_function(aes(color="Lognormal, sdlog=0.5"),
                fun=dlnorm,
                args=list(meanlog = log(0.2), sdlog = 0.5))+
  geom_function(aes(color="Lognormal, sdlog=2"),
                fun=dlnorm,
                args=list(meanlog = log(0.2), sdlog = 2))+
  geom_function(aes(color="Uniform"),
                fun=dunif,
                args=list(min=0,max=1))+
  ggtitle("Priors for r with median of 0.2")+
  labs(x="r",y="Probability",color="")
ggplot(data.frame(K=c(1000,100000)),aes(x=K))+
  geom_function(aes(color="Lognormal\nmean=50000, sdlog=2"),
                fun=dlnorm,
                args=list(meanlog = log(50000), sdlog = 2))+
  geom_function(aes(color="Uniform on log"),
                fun=dlogunif,
                args=list(minval=1000,maxval=100000,logTrans=FALSE))+
  geom_function(aes(color="Uniform"),
                fun=dunif,
                args=list(min=1000,max=100000))+
  ggtitle("Priors for K, lnorm has median of 50000")+
  labs(x="K",y="Probability",color="")
ggplot(data.frame(phi=c(0,2)),aes(x=phi))+
  geom_function(aes(color="Lognormal\nmean=1, sdlog=0.1"),
                fun=dlnorm,
                args=list(meanlog = log(1), sdlog = 0.1))+
  geom_function(aes(color="Lognormal\nmean=1, sdlog=0.5"),
                fun=dlnorm,
                args=list(meanlog = log(1), sdlog = 0.2))+
  geom_function(aes(color="Lognormal\nmean=1, sdlog=0.2"),
                fun=dlnorm,
                args=list(meanlog = log(1), sdlog = 0.5))+
  ggtitle("Priors for starting biomass ratio B0/K with median of 1")+
  labs(x="phi=Bo/K",y="Probability",color="")

```

For small values of log-sd, the log-sd is very similar to the CV which is mean(x)/sd(x), so they are sometimes used interchangeably. 

### Using JABBA

First, install JABBA and JAGS

```{r}
# Installation
# install.packages(devtools)
# devtools:: install_github("jabbamodel/JABBA")
```

Original South Atlantic Swordfish example
Winker et al. (2018). JABBA: Just Another Biomass Assessment

```{r}
library(JABBA)
swos <- iccat$swos

assessment <- "SWOSiccat"
scenario <- "Schaefer"

File <- getwd() #"C:/Users/Elizabeth A Babcock/Box/RSM 612/RSM 612 lectures" # LINK to your folder of choice here
assessment<-"SWO"
output.dir = file.path(File,assessment)
dir.create(output.dir,showWarnings = FALSE)
```

### Compile JABBA JAGS model and input object

```{r}
jbswos <- build_jabba(catch=swos$catch,  #vector of catches
                      cpue=swos$cpue,  #data frame of CPUE by fleet
                      se=swos$se,  #input standard deviations.
                      assessment=assessment, #name of assessment
                      scenario = scenario, #scenario name
                      model.type = "Schaefer",  #Schaefer is the logistic model
                      r.prior=c(0.42,0.37), #mean and log-sd for lognormal
                      K.prior = c(250000,1), #mean and log-sd for lognormal
                      psi.prior = c(1,0.25), #mean and log-sd for lognormal
                      sigma.est = TRUE,  # observation error by fleet
                      igamma=c(4,0.01), # default process error
                      P_bound = c(0.02,1.1),
                      verbose=FALSE, #Bounds of B/K
                      sigmaobs_bound = 0.1, #upper limit of observation error
                      sigmaproc_bound = 0.2) #upper limit of process error)
```

Plot indices

```{r}
jbplot_indices(jbswos)
```

fit JABBA

This is running a model in JAGS so it takes some time

```{r}
fit.swos <- fit_jabba(jbswos,  #the build_jabba object
                      save.jabba=TRUE,  #TO save the directory
                      output.dir=output.dir,
                      ni = 50000,  #ni,nt,nb and nc are MCMC settings
                      nt = 10,
                      nb = 5000,
                      nc = 2)
```

Plots to see if the MCMC converged

```{r}
jbplot_mcmc(fit.swos)
jbplot_ppdist(fit.swos)  #Posterior and prior distributions
jbplot_PPC(fit.swos)  #Posterior predictive checks
```


Make some plots to see if the fit is good

```{r}
jbplot_catch(fit.swos)  #catches
jbplot_cpuefits(fit.swos)  #Fits to CPUE indices
jbplot_logfits(fit.swos)   #Fits to CPUE indices in log scale
jbplot_residuals(fit.swos)  # Residuals
jbplot_runstest(fit.swos)  # A test of whether there are patterns in residuals
```

Plots to see the status of the population

```{r}
jbplot_ensemble(fit.swos,subplots=1:2)
jbplot_kobe(fit.swos)
jbplot_spphase(fit.swos)
```


### Projections

 ICCAT Style projections, based on a TAC: total allowable catch

```{r}
Ccur = 10056
TACs = seq(10000,18000,2000)
fw.swos= fw_jabba(fit.swos,nyears=10,imp.yr=4,
                  initial = Ccur,
                  imp.values = TACs,
                  quant="Catch",type="abs",
                  nsq=3,stochastic = T,AR1=F)
jbplot_ensemble(fw.swos,legendcex = 0.4,xlim=c(2015,2025),subplots = c(1:4),add=T)
```

### Hindcast cross-validation and retrospective analysis

```{r}
hc1 <- hindcast_jabba(jbswos,fit.swos,peels=1:5)
 # Show Retrospective Pattern
jbplot_retro(hc1)
jbplot_hcxval(hc1)
```

### Pella-Tomlinson model

```{r}
scenario<-"Pella"
jbswos <- build_jabba(catch=swos$catch,  #vector of catches
                      cpue=swos$cpue,  #data frame of CPUE by fleet
                      se=swos$se,  #input standard deviations.
                      assessment=assessment, #name of assessment
                      scenario = scenario, #scenario name
                      model.type = "Pella",  #Pella is an assymetric yield curve,
                      BmsyK = 0.4, #Location of Bmsy/K
                      r.prior=c(0.42,0.37), #mean and log-sd for lognormal
                      K.prior = c(250000,1), #mean and log-sd for lognormal
                      psi.prior = c(1,0.25), #mean and log-sd for lognormal
                      sigma.est = TRUE,  # observation error by fleet
                      igamma=c(4,0.01), # default process error
                      P_bound = c(0.02,1.1), #Bounds of B/K
                      sigmaobs_bound = 0.1, #upper limit of observation error
                      sigmaproc_bound = 0.2) #upper limit of process error)
fit.swos2 <- fit_jabba(jbswos,  #the build_jabba object
                      save.jabba=TRUE,  #TO save the directory
                      output.dir=output.dir,
                      ni = 50000,  #ni,nt,nb and nc are MCMC settings
                      nt = 10,
                      nb = 5000,
                      nc = 2)

#save.image()
```

Plot yield curve with status

```{r}
jbplot_spdyn(fit.swos2)
```

Compare runs

```{r}
jbplot_ensemble(list(fit.swos,fit.swos2),subplots=1:2)
jbplot_kobe(fit.swos2)

```