---
title: "New Runs"
author: "Beth Babcock"
date: "`r Sys.Date()`"
output: html_document
---

### Using JABBA

First, install JABBA and JAGS

```{r}
# Installation
# install.packages(devtools)
# devtools:: install_github("jabbamodel/JABBA")
```

Original South Atlantic Swordfish example
Winker et al. (2018). JABBA: Just Another Biomass Assessment

```{r}
library(JABBA)
swos <- iccat$swos

assessment <- "SWOSiccat"
scenario <- "Schaefer"

File <- getwd() #"C:/Users/Elizabeth A Babcock/Box/RSM 612/RSM 612 lectures" # LINK to your folder of choice here
assessment<-"SWO"
output.dir = file.path(File,assessment)
dir.create(output.dir,showWarnings = FALSE)
```

### Compile JABBA JAGS model and input object

```{r}
jbswos <- build_jabba(catch=swos$catch,  #vector of catches
                      cpue=swos$cpue,  #data frame of CPUE by fleet
                      se=swos$se,  #input standard deviations.
                      assessment=assessment, #name of assessment
                      scenario = scenario, #scenario name
                      model.type = "Schaefer",  #Schaefer is the logistic model
                      r.prior=c(0.42,1), #mean and log-sd for lognormal
                      K.prior = c(250000,1), #mean and log-sd for lognormal
                      psi.prior = c(1,0.25), #mean and log-sd for lognormal
                      sigma.est = TRUE,  # observation error by fleet
                      igamma=c(4,0.01), # default process error
                      P_bound = c(0.02,1.1),
                      verbose=FALSE, #Bounds of B/K
                      sigmaobs_bound = 0.1, #upper limit of observation error
                      sigmaproc_bound = 0.05) #upper limit of process error)
```

Plot indices

```{r}
jbplot_indices(jbswos)
```

fit JABBA

This is running a model in JAGS so it takes some time

```{r}
fit.swos <- fit_jabba(jbswos,  #the build_jabba object
                      save.jabba=TRUE,  #TO save the directory
                      output.dir=output.dir,
                      ni = 500000,  #ni,nt,nb and nc are MCMC settings
                      nt = 100,
                      nb = 5000,
                      nc = 2)
```

Plots to see if the MCMC converged

```{r}
jbplot_mcmc(fit.swos)
jbplot_ppdist(fit.swos)  #Posterior and prior distributions
jbplot_PPC(fit.swos)  #Posterior predictive checks
```


Make some plots to see if the fit is good

```{r}
jbplot_catch(fit.swos)  #catches
jbplot_cpuefits(fit.swos)  #Fits to CPUE indices
jbplot_logfits(fit.swos)   #Fits to CPUE indices in log scale
jbplot_residuals(fit.swos)  # Residuals
jbplot_runstest(fit.swos)  # A test of whether there are patterns in residuals
```

Plots to see the status of the population

```{r}
jbplot_ensemble(fit.swos,subplots=1:2)
jbplot_kobe(fit.swos)
jbplot_spphase(fit.swos)
```


### Projections

 ICCAT Style projections, based on a TAC: total allowable catch

```{r}
Ccur = 10056
TACs = seq(10000,18000,2000)
fw.swos= fw_jabba(fit.swos,nyears=10,imp.yr=4,
                  initial = Ccur,
                  imp.values = TACs,
                  quant="Catch",type="abs",
                  nsq=3,stochastic = T,AR1=F)
jbplot_ensemble(fw.swos,legendcex = 0.4,xlim=c(2015,2025),subplots = c(1:4),add=T)
```

### Hindcast cross-validation and retrospective analysis

```{r}
hc1 <- hindcast_jabba(jbswos,fit.swos,peels=1:5)
 # Show Retrospective Pattern
jbplot_retro(hc1)
jbplot_hcxval(hc1)
```
